---
title: "LW Chapter 3"
author: "mht"
date: "October 5, 2014"
output: html_document
---


```{r setup}
#setwd("~/Repos/psych201s/practicums/")
library(rwebppl)
rm(list=ls())
```

### 3.0 Warm up

We saw at the end of the last tutorial (bayes-rule) that we can write down a model in WebPPL like this.

```{r}
myPosterior <- "
var model = function(){
  var p = sample(Uniform( {a:0, b:1} )) // prior on coin weights
  observe(
  { data : 15,   // observed data: numbers of heads
    link : Binomial( {n:20, p: p} ) } // linking function is a Binomial 
  )
  return p
}
"

posterior <- webppl(
  program_code = myPosterior,
  model_var = "model",
  inference_opts = list(method = "rejection", samples = 10000),
  output_format = "samples",
  packages = c("./utils")
)
```

Sometimes, we will want to try out different numbers in our models (e.g., what if we observed 16 heads instead of 15 heads?), and it would be a pain to copy and paste the model everytime we wanted to try different numbers. Fortunately, RWebPPL allows us to pass data from R to WebPPL.

```{r}
modelWithNoData <- '
// data is currently passed as a list
// when there is just one data point, grab the 0th element of the list
var number_heads = data_from_R["number_heads"][0]
var number_flips = data_from_R["number_flips"][0]

var model = function(){
  var p = sample(Uniform( {a:0, b:1} )) // prior on coin weights
  observe({ 
    data : number_heads,   // observed data: numbers of heads
    link : Binomial( {n:number_flips, p: p} ) } // linking function is a Binomial 
  )
  return p
}
'

data_to_webppl <- list(number_heads = 15, number_flips = 20)

posterior <- webppl(
  program_code = modelWithNoData,
  model_var = "model",
  data_var = "data_from_R",
  data = data_to_webppl,
  inference_opts = list(method = "rejection", samples = 10000),
  output_format = "samples",
  packages = c("./utils")
)

```

Note that `data_from_R` is not defined inside of the webppl program `modelWithNoData`. We are passing it to the webppl program via the `data` argument to `webppl(...)` (the data is in a list in R that we've defined as `data_to_webppl`). We also need to tell webppl what to expect our data to be called inside of the webppl program. This is done with the `data_var` argument.


Take another look at the arguments to `webppl(...)` call. What is each of them doing? 


In this chapter, we will be running a lot of models with the same basic format. In R, we can make another function that abstracts much of this code into a function called `runModel`

```{r}
runModel <- function(model, data_to_webppl, 
                    numSamples = 10000, method = "rejection") {
  # The following command calls WebPPL with specific options.
  wp <- webppl(
    program_code = model,
    data = data_to_webppl,
    data_var = "observed_data", # we will always refer inside WebPPL to the data passed from R: "observed_data"
    inference_opts = list(method = method, samples = numSamples),
    model_var = "model", # we will always call our model to do inference over: "model"
    output_format = "samples",
    packages = c("./utils")
    )
}
```

Here's the model from above re-written using the expected parameter names, using the helper function. 

```{r}
modelWithNoData <- '
// data is currently passed as a list
// when there is just one data point, grab the 0th element of the list
var number_heads = observed_data["number_heads"][0]
var number_flips = observed_data["number_flips"][0]

var model = function(){
  var p = sample(Uniform( {a:0, b:1} )) // prior on coin weights
  observe({ 
    data : number_heads,   // observed data: numbers of heads
    link : Binomial( {n:number_flips, p: p} ) } // linking function is a Binomial 
  )
  return p
}
'
data_to_webppl = list(number_heads = 15, number_flips = 20)
rs <- runModel(modelWithNoData, data_to_webppl)

```


### 3.1 Inferring a rate


```{r}
myFirstModel <- '
// Unpack data
var k = observed_data[0]["k"] // number of heads
var n = observed_data[0]["n"] // number of flips

var model = function() {
   var p = uniform( {a:0, b:1} )    // sample from a uniform
   observe({
      data : k,           // Observed k number of Heads
      link: Binomial( {p : p, n: n }) // assuming a Binomial distribution
   })
  // var predictive = sample(Binomial({p : p, n : n}))
  return {p: p}
}
'

data_to_webppl = data.frame(k = 15, n = 20)

posterior <- runModel(myFirstModel, data_to_webppl) 

ggplot(posterior, aes(x = p)) +
  geom_histogram()
```


#3.1.1
Hint: remember that the probability of an interval is area under the curve


#3.1.2


#3.1.3



```{r, echo=FALSE}
# Enter new k & n values here:
observed_data <- data.frame(k = ..., n = ...)

runModel(myFirstModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)

```


#3.1.4

Hint: we used 10000 samples above

```{r}
observed_data <- data.frame(k = 5, n = 10)
runModel(myFirstModel, observed_data, numSamples = ...) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```

```{r}
observed_data <- data.frame(k = 50, n = 100) 
runModel(myFirstModel, observed_data, numSamples = ...) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```

#3.1.5

```{r}
observed_data <- data.frame(k = ..., n = ...) 
runModel(myFirstModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)

```

#3.1.6

```{r}
observed_data <- data.frame(k = ..., n = ...) 
runModel(myFirstModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```
  
### 3.2 Difference between two rates


```{r}
# k1 = number of heads for coin 1
# k2 = number of heads for coin 2
# n1 = number of coin flips for coin 1
# n2 = number of coin flips for coin 2
differenceInWeightsModel <- '
// Unpack data
var k1 = observed_data["k1"]
var k2 = observed_data["k2"]
var n1 = observed_data["n1"]
var n2 = observed_data["n2"]

var model = function() {
  // sample two rates from uniform priors
  var p1 = uniform({a: 0, b: 1})
  var p2 = uniform({a: 0, b: 1})

  // Take first observation into account
  observe({
    data : k1,          
    link: Binomial( {p : p1, n: n1 }) 
  }) 

  // Take second observation into account
  observe({
    data : k2,          
    link: Binomial( {p : p2, n: n2 }) 
  }) 

  // Return the difference
  return {delta : p1 - p2}
}
'
```

So far, we've been using the method of Rejection Sampling, which corresponds to the very simple algorithm we saw in `bayes-rule.Rmd`. Though intuitive and mathematically correct, rejection sampling is inefficient. It is inefficient because we have to accurately predict the data we've observed... if we don't match it exactly, we throw away the sample and start over. As our data set grows bigger, the exact values in it become more unlikely (due to the *AND* rule, or multiplication rule, of probability). You have to predict this data point AND this data point AND this data point, etc...

Researchers in computer science have known this and have developed other so called "inference algorithms". That is, techniques for doing Bayesian inference. That are more efficient. The details of these are beyond the scope of this class, but you should be away that they exist. Throughout these practica, we will may change the `inference_opts` argument to use an algorithm that is more efficient. 


```{r}

runModel32 <- function(model, data_to_webppl, 
                    numSamples = 5000) {
  wp <- webppl(
    program_code = model,
    data = data_to_webppl,
    data_var = "observed_data", 
       inference_opts = list(method="MCMC", 
                             kernel = 
                               list(HMC = 
                                      list(steps = 10,
                                           stepSize = 0.05)),
                          samples = numSamples, 
                          burn = numSamples/2,
                          verbose = TRUE),
    model_var = "model",
    output_format = "samples",
    packages = c("./utils")
    )
}

```

```{r}
observed_data = list(k1 = 5, n1 = 10, k2 = 7, n2 = 10)

posterior <- runModel32(differenceInWeightsModel, observed_data)

ggplot(wp.samp, aes(x = delta)) +
  geom_histogram(binwidth = .025)
```

# 3.2.1

Run with first set of params:

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel32(differenceInWeightsModel, observed_data) %>%
  ggplot(aes(x = delta)) +
  geom_histogram(binwidth = .025)
```

With second set of params:

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel32(differenceInWeightsModel, observed_data) %>%
  ggplot(aes(x = delta)) +
  geom_histogram(binwidth = .025)
```

# 3.2.2

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel32(differenceInWeightsModel, observed_data) %>%
  ggplot(aes(x = delta)) +
  geom_histogram(binwidth = .025)
```

# 3.2.3



### 3.3 Inferring a common rate

```{r}
commonRateModel <- '
// Unpack data
var k1 = observed_data["k1"]
var k2 = observed_data["k2"]
var n1 = observed_data["n1"]
var n2 = observed_data["n2"]

var model = function() {
  // Sample rate from uniform distribution
  var p = uniform( {a:0, b:1} )
  
  // account for first data point using p
  observe({
    data : k1,           // Observed k number of Heads
    link: Binomial( {p : p, n: n1 }) // assuming a Binomial distribution
  }) 

  // account for second data point also using p
  observe({
    data : k2,           // Observed k number of Heads
    link: Binomial( {p : p, n: n2 }) // assuming a Binomial distribution
  }) 

  return {p : p}
}
'
```

Run the model, example:

```{r}
observed_data = list(k1 = 5, n1 = 10, k2 = 7, n2 = 10)
posterior <- runModel32(commonRateModel, observed_data)

ggplot(posterior, aes(x = p)) +
  geom_histogram(binwidth = .025)
```

# 3.3.1

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```

# 3.3.2

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```

# 3.3.3

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = .025)
```
