---
title: "LW Chapter 3"
author: "mht"
date: "October 5, 2014"
output: html_document
---

### 3.1 Inferring a rate

```{r setup}
setwd("/Users/mht/Documents/learning/class/teaching/16.3/practicums/")
library(rwebppl)
library(ggmcmc)
#library(gridExtra)
rm(list=ls())
```

```{r}
model1 <- '
// Unpack data
var k = observed_data["k"] // number of heads
var n = observed_data["n"] // number of flips

var myFirstModel = function() {
   var p = uniform( {a:0, b:1} )    // sample a coin weight from a uniform distribution
   observe({
      data : k,           // Observed k number of Heads
      link: Binomial( {p : p, n: n }) // assuming a Binomial distribution
   }) 
   return {nextFlip: flip(p)}
}
'
```


```{r}
## data
# k = number of heads
# n = number of coin flips
observed_data <- list(k = 5, n = 10) # packaging data to pass to WebPPL

# The following command calls WebPPL with specific options.
wp <- webppl(
  program_code = model1,
  data = observed_data,
  data_var = "observed_data",
  inference_opts = list(method="MCMC", samples = 10000, burn = 5000),
  model_var = "myFirstModel",
#  ggmcmc = TRUE,
  packages = c("./utils")
  )

summary(wp)

# Now let's plot a histogram for theta. 
ggs_histogram(wp)+xlim(0,1)
# some credible interval
ggs_caterpillar(wp)+xlim(0,1)

# Now the values for the monitored parameters are in the "samples" object, 
# ready for inspection.

# The commands below are useful for a quick overview:
print(samples)  # a rough summary
plot(samples)   # a visual representation
traceplot(samples) # traceplot (press <enter> repeatedly to see the chains)

#more info on what is returned:
summary(samples)
summary(samples$BUGSoutput)

chain <- 1
samples$BUGSoutput$sims.array[1:15,chain,]# array: element, chain, column (theta/deviance) 
```

Writing a model as a string in R is convenient, but it is also convenient to see syntax-highlighting (like you get syntax highlighting for R). To get this, we can just write out model as a .wppl file in our folder. You can open the file in RMarkdown so you can view it and edit it here. In the bottom-right of the editor, you can change the language to JavaScript to get some helpful comments. (Note there are a few differences between JavaScript and WebPPL so all of the comments may not apply).


```{r}
wp <- webppl(
  program_file = "webppl_models/myFirstModel.wppl",
  packages = c("./utils"),
  data = observed_data,
  data_var = "observed_data",
  inference_opts = list(method="MCMC", samples = 10000, burn = 5000),
  model_var = "myFirstModel",
  ggmcmc = TRUE
  )
```


#3.1.1
Based on visual inspection, my estimate that the probability is higher than 0.4 and lower than 0.6 is roughly 50%. I did this by drawing a line at 0.4 and 0.6 and estimating the area with respect to the entire area.
#3.1.2
It seems to be about twice as likely that theta is equal to 0.5 than 0.7. I looked at the relative height of the two points.

#3.1.3
```{r, echo=FALSE}
k <- 10 # number of heads
n <- 20 # number of samples


samples <- jags(data, inits=myinits, parameters,
   			 model.file =f, n.chains=2, n.iter=20000, 
         n.burnin=1, n.thin=1, DIC=T)
# Collect posterior samples across all chains:
posterior <- data.frame(samples$BUGSoutput$sims.list$theta)
names(posterior)<-c('theta')
# Now let's plot a histogram for theta. 

ggplot(data=posterior,aes(x=theta))+
 # geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2)+
  theme_solarized()+
  ggtitle('3.1.3: k=50, n=100')+
  xlim(0,1)
```
This is much narrower than before.

#3.1.4
```{r}
k <- 5 # number of heads
n <- 10 # number of samples


samples <- jags(data, 
                inits=myinits, 
                parameters,
     		 model.file =f, 
     		 n.chains=2, 
     		 n.iter=200000, 
         n.burnin=1, 
     		 n.thin=1, 
     		 DIC=T)
# Collect posterior samples across all chains:
posterior <- data.frame(samples$BUGSoutput$sims.list$theta)
names(posterior)<-c('theta')
# Now let's plot a histogram for theta. 
ggplot(data=posterior,aes(x=theta))+
  geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  theme_bw()+
  ggtitle('3.1.4a: k=5, n=10, n.iter=10x')+
  xlim(0,1)



k <- 50 # number of heads
n <- 100 # number of samples


samples <- jags(data, inits=myinits, parameters,
       	 model.file =f, n.chains=2, n.iter=200000, 
         n.burnin=1, n.thin=1, DIC=T)
# Collect posterior samples across all chains:
posterior <- data.frame(samples$BUGSoutput$sims.list$theta)
names(posterior)<-c('theta')
# Now let's plot a histogram for theta. 
ggplot(data=posterior,aes(x=theta))+
  geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  theme_bw()+
  ggtitle('3.1.4b: k=50, n=100, n.iter=10x')+
  xlim(0,1)
```

The width is controlled by the evidence (k & n). The smoothness is controlled by the number of smaples (n.iter).

#3.1.5

```{r}
k <- 99 # number of heads
n <- 100 # number of samples


samples <- jags(data, inits=myinits, parameters,
       	 model.file =f, n.chains=2, n.iter=200000, 
         n.burnin=1, n.thin=1, DIC=T)
# Collect posterior samples across all chains:
posterior <- data.frame(samples$BUGSoutput$sims.list$theta)
names(posterior)<-c('theta')
# Now let's plot a histogram for theta. 
ggplot(data=posterior,aes(x=theta))+
  geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  theme_bw()+
    ggtitle('3.1.5')+
  xlim(0,1)
```

Non symmetric, very close to 1.

```{r}
k <- 0 # number of heads
n <- 1 # number of samples


samples <- jags(data, inits=myinits, parameters,
          model.file =f, n.chains=2, n.iter=200000, 
         n.burnin=1, n.thin=1, DIC=T)
# Collect posterior samples across all chains:
posterior <- data.frame(samples$BUGSoutput$sims.list$theta)
names(posterior)<-c('theta')
# Now let's plot a histogram for theta. 
ggplot(data=posterior,aes(x=theta))+
  geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  theme_bw()+
  ggtitle('3.1.6')+
  xlim(0,1)
```
Even with very little evidence, different from the prior. 

### 3.2 Difference between two rates

```{r}

k1 <- 3
k2 <- 11
n1 <- 20
n2 <- 20

cat('# Difference Between Two Rates
model{
   # Observed Counts
   k1 ~ dbin(theta1,n1)
   k2 ~ dbin(theta2,n2)
   # Prior on Rates
   theta1 ~ dbeta(1,1)
   theta2 ~ dbeta(1,1)
   # Difference Between Rates
   delta <- theta1-theta2
}',file={f<-tempfile()})

data <- list("k1", "k2", "n1", "n2") # to be passed on to JAGS
myinits <-  list(
  list(theta1 = 0.1, theta2 = 0.9))

# parameters to be monitored:	
parameters <- c("delta", "theta1", "theta2")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file =f, n.chains=1, n.iter=10000, 
         n.burnin=1, n.thin=1, DIC=T)
# Now the values for the monitored parameters are in the "samples" object, 
# ready for inspection.

post_delta <- data.frame(samples$BUGSoutput$sims.list$delta)
names(post_delta)<-c('delta')
ggplot(data=post_delta,aes(x=delta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='blue')+
  theme_bw()+
  ggtitle('3.2')+
  xlim(-1,1)

post_theta <- data.frame(theta1=samples$BUGSoutput$sims.list$theta1,
                         theta2=samples$BUGSoutput$sims.list$theta2)

d.t<- gather(post_theta,key,val)
ggplot(data=d.t,aes(x=val,fill=key))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,alpha=0.7)+
  theme_bw()+
  xlim(0,1)



# mean of delta:
mean(post_delta$delta)
# median of delta:
median(post_delta$delta)
# mode of delta, estimated from the "density" smoother:
density(post_delta$delta)$x[which(density(post_delta$delta)$y==max(density(post_delta$delta)$y))]
# 95% credible interval for delta:
quantile(post_delta$delta, c(.025,.975))
```


```{r}

k1 <- 8
k2 <- 7
n1 <- 10
n2 <- 10
samples <- jags(data, inits=myinits, parameters,
   			 model.file =f, n.chains=1, n.iter=10000, 
         n.burnin=1, n.thin=1, DIC=T)
# Now the values for the monitored parameters are in the "samples" object, 
# ready for inspection.

post_delta <- data.frame(samples$BUGSoutput$sims.list$delta)
names(post_delta)<-c('delta')
ggplot(data=post_delta,aes(x=delta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='blue')+
  theme_bw()+
  ggtitle('3.2.1a: k1=8, n1=10; k2=7; n2=10')+
  xlim(-1,1)

k1 <- 80
k2 <- 70
n1 <- 100
n2 <- 100
samples <- jags(data, inits=myinits, parameters,
     		 model.file ="Rate_2.txt", n.chains=1, n.iter=10000, 
         n.burnin=1, n.thin=1, DIC=T)
# Now the values for the monitored parameters are in the "samples" object, 
# ready for inspection.

post_delta <- data.frame(samples$BUGSoutput$sims.list$delta)
names(post_delta)<-c('delta')
ggplot(data=post_delta,aes(x=delta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='blue')+
  theme_bw()+
  ggtitle('3.2.1b: k1=80, n1=100; k2=70; n2=100')+
  xlim(-1,1)

```


```{r}
k1 <- 0
k2 <- 0
n1 <- 1
n2 <- 5
samples <- jags(data, inits=myinits, parameters,
     		 model.file =f, n.chains=1, n.iter=10000, 
         n.burnin=1, n.thin=1, DIC=T)
# Now the values for the monitored parameters are in the "samples" object, 
# ready for inspection.

post_delta <- data.frame(samples$BUGSoutput$sims.list$delta)
names(post_delta)<-c('delta')
ggplot(data=post_delta,aes(x=delta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='blue')+
  theme_bw()+
  ggtitle('3.2.2: k1=0, n1=1; k2=0; n2=5')+
  xlim(-1,1)
```

We're more certain that theta2 is 0. Theta1 is a prior uniformly distributed between 0 and 1. Hence, we are more certain than Theta2 is less than Theta 1 --> delta > 0


# 3.2.3

Point estimates - unimodal
Credible interval - unimodal


### 3.3 Inferring a common rate

```{r}

rm(list=ls()) 

k1 <- 14
k2 <- 16
n1 <- 20
n2 <- 20

cat('# Inferring a Common Rate
model{  
   # Observed Counts
   k1 ~ dbin(theta,n1)
   k2 ~ dbin(theta,n2)
   # Prior on Single Rate Theta
   theta ~ dbeta(1,1)  
}',file={f<-tempfile()})

data <- list("k1", "k2", "n1", "n2") # to be passed on to JAGS
myinits <-  list(
  list(theta = 0.5))

# parameters to be monitored:	
parameters <- c("theta")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file =f, n.chains=1, n.iter=1000, 
         n.burnin=1, n.thin=1, DIC=T)

post_theta <- data.frame(samples$BUGSoutput$sims.list$theta)
names(post_theta)<-c('theta')
ggplot(data=post_theta,aes(x=theta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='green')+
  theme_bw()+
  ggtitle('3.3.1: k1=14, n1=20; k2=16; n2=20')+
  xlim(0,1)

mean(post_theta$theta)
median(post_theta$theta)
quantile(post_theta$theta, c(.025,.975))
```

# 3.3.1
Seems like the 95% interval would be best here.

```{r}

k1 <- 0
k2 <- 10
n1 <- 10
n2 <- 10

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file ="Rate_3.txt", n.chains=1, n.iter=1000, 
         n.burnin=1, n.thin=1, DIC=T)

post_theta <- data.frame(samples$BUGSoutput$sims.list$theta)
names(post_theta)<-c('theta')
quartz()
ggplot(data=post_theta,aes(x=theta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='green')+
  theme_bw()+
  ggtitle('3.3.2: k1=0, n1=10; k2=10; n2=10')+
  xlim(0,1)

mean(post_theta$theta)
median(post_theta$theta)
quantile(post_theta$theta, c(.025,.975))
```

# 3.3.2
No; might need a more sophisticated model (more than 1 latent cause).

```{r}

k1 <- 7
k2 <- 3
n1 <- 10
n2 <- 10

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
   			 model.file ="Rate_3.txt", n.chains=1, n.iter=1000, 
         n.burnin=1, n.thin=1, DIC=T)

post_theta <- data.frame(samples$BUGSoutput$sims.list$theta)
names(post_theta)<-c('theta')

quartz()
ggplot(data=post_theta,aes(x=theta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='green')+
  theme_bw()+
  ggtitle('3.3.3a: k1=7, n1=10; k2=3; n2=10')+
  xlim(0,1)

mean(post_theta$theta)
median(post_theta$theta)
quantile(post_theta$theta, c(.025,.975))


k1 <- 5
k2 <- 5
n1 <- 10
n2 <- 10

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
     		 model.file ="Rate_3.txt", n.chains=1, n.iter=20000, 
         n.burnin=1, n.thin=1, DIC=T)

post_theta <- data.frame(samples$BUGSoutput$sims.list$theta)
names(post_theta)<-c('theta')

quartz()
ggplot(data=post_theta,aes(x=theta))+
  #geom_histogram(aes(y=..count../sum(..count..)),binwidth=0.02)+
  geom_density(size=2,colour='green')+
  theme_bw()+
  ggtitle('3.3.3b: k1=5, n1=10; k2=5; n2=10')+
  xlim(0,1)

mean(post_theta$theta)
median(post_theta$theta)
quantile(post_theta$theta, c(.025,.975))
```


Why isn't it a narrower distribution in the second case?


### 3.4 Prior and posterior prediction

```{r}
k <- 1
n <- 15
# Uncomment for Trompetter Data
# k <- 24
# n <- 121

data <- list("k", "n") # to be passed on to JAGS
myinits <-  list(list(theta = 0.5, thetaprior = 0.5))

# parameters to be monitored:	
parameters <- c("theta", "thetaprior", "postpredk", "priorpredk")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file ="Rate_4.txt", n.chains=1, n.iter=5000, 
         n.burnin=1, n.thin=1, DIC=T)

df<- data.frame(samples$BUGSoutput$sims.list$theta)
names(df)[1]<-'theta'
df$thetaprior <- samples$BUGSoutput$sims.list$thetaprior 
df$postpredk  <- samples$BUGSoutput$sims.list$postpredk 
df$priorpredk <- samples$BUGSoutput$sims.list$priorpredk
mlt<-melt(df)
mlt$param = NA
mlt$dstr = NA
mlt[mlt$variable%in%c('theta','thetaprior'),]$param = 'theta'
mlt[mlt$variable%in%c('postpredk','priorpredk'),]$param = 'k'
mlt[mlt$variable%in%c('priorpredk','thetaprior'),]$dstr = 'prior'
mlt[mlt$variable%in%c('theta','postpredk'),]$dstr = 'post'

ggplot(mlt, aes(x=value, fill=dstr)) + 
  facet_wrap(~param,scales='free') + 
  geom_histogram(data = subset(mlt, param == 'k'),position='dodge',binwidth=1) +
  geom_density(data = subset(mlt, param == 'theta'))+
  theme_bw()
```

# 3.4.2

```{r}
k <- 1
n <- 15
# Uncomment for Trompetter Data
# k <- 24
# n <- 121

data <- list("k", "n") # to be passed on to JAGS
myinits <-  list(list(theta = 0.5, thetaprior = 0.5))

# parameters to be monitored:  
parameters <- c("theta", "thetaprior", "postpredk", "priorpredk")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file ="Rate_4_342.txt", n.chains=1, n.iter=5000, 
         n.burnin=1, n.thin=1, DIC=T)

df<- data.frame(samples$BUGSoutput$sims.list$theta)
names(df)[1]<-'theta'
df$thetaprior <- samples$BUGSoutput$sims.list$thetaprior 
df$postpredk  <- samples$BUGSoutput$sims.list$postpredk 
df$priorpredk <- samples$BUGSoutput$sims.list$priorpredk
mlt<-melt(df)
mlt$param = NA
mlt$dstr = NA
mlt[mlt$variable%in%c('theta','thetaprior'),]$param = 'theta'
mlt[mlt$variable%in%c('postpredk','priorpredk'),]$param = 'k'
mlt[mlt$variable%in%c('priorpredk','thetaprior'),]$dstr = 'prior'
mlt[mlt$variable%in%c('theta','postpredk'),]$dstr = 'post'

ggplot(mlt, aes(x=value, fill=dstr)) + 
  facet_wrap(~param,scales='free') + 
  geom_histogram(data = subset(mlt, param == 'k'),position='dodge',binwidth=1) +
  geom_density(data = subset(mlt, param == 'theta'))+
  theme_bw()
```


# 3.4.3

```{r}
k <- 1
n <- 100
m <- 10
# Uncomment for Trompetter Data
# k <- 24
# n <- 121

data <- list("k", "n", "m") # to be passed on to JAGS
myinits <-  list(list(theta = 0.5, thetaprior = 0.5))

# parameters to be monitored:  
parameters <- c("theta", "thetaprior", "postpredk", "priorpredk")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
   			 model.file ="Rate_4_343.txt", n.chains=1, n.iter=5000, 
         n.burnin=1, n.thin=1, DIC=T)

df<- data.frame(samples$BUGSoutput$sims.list$theta)
names(df)[1]<-'theta'
df$thetaprior <- samples$BUGSoutput$sims.list$thetaprior 
df$postpredk  <- samples$BUGSoutput$sims.list$postpredk 
df$priorpredk <- samples$BUGSoutput$sims.list$priorpredk
mlt<-melt(df)
mlt$param = NA
mlt$dstr = NA
mlt[mlt$variable%in%c('theta','thetaprior'),]$param = 'theta'
mlt[mlt$variable%in%c('postpredk','priorpredk'),]$param = 'k'
mlt[mlt$variable%in%c('priorpredk','thetaprior'),]$dstr = 'prior'
mlt[mlt$variable%in%c('theta','postpredk'),]$dstr = 'post'

ggplot(mlt, aes(x=value, fill=dstr)) + 
  facet_wrap(~param,scales='free') + 
  geom_histogram(data = subset(mlt, param == 'k'),position='dodge',binwidth=1) +
  geom_density(data = subset(mlt, param == 'theta'))+
  theme_bw()
```

# 3.4.2

```{r}
#k <- 1
#n <- 100
#m <- 10
# Uncomment for Trompetter Data
k <- 24
n <- 121

data <- list("k", "n") # to be passed on to JAGS
myinits <-  list(list(theta = 0.5, thetaprior = 0.5))

# parameters to be monitored:  
parameters <- c("theta", "thetaprior", "postpredk", "priorpredk")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
     		 model.file ="Rate_4.txt", n.chains=1, n.iter=5000, 
         n.burnin=1, n.thin=1, DIC=T)

df<- data.frame(samples$BUGSoutput$sims.list$theta)
names(df)[1]<-'theta'
df$thetaprior <- samples$BUGSoutput$sims.list$thetaprior 
df$postpredk  <- samples$BUGSoutput$sims.list$postpredk 
df$priorpredk <- samples$BUGSoutput$sims.list$priorpredk
mlt<-melt(df)
mlt$param = NA
mlt$dstr = NA
mlt[mlt$variable%in%c('theta','thetaprior'),]$param = 'theta'
mlt[mlt$variable%in%c('postpredk','priorpredk'),]$param = 'k'
mlt[mlt$variable%in%c('priorpredk','thetaprior'),]$dstr = 'prior'
mlt[mlt$variable%in%c('theta','postpredk'),]$dstr = 'post'

ggplot(mlt, aes(x=value, fill=dstr)) + 
  facet_wrap(~param,scales='free') + 
  geom_histogram(data = subset(mlt, param == 'k'),position='dodge',binwidth=1) +
  geom_density(data = subset(mlt, param == 'theta'))+
  theme_bw()
```



### 3.5 Posterior prediction


```{r}

k1 <- 0
k2 <- 10
n1 <- 10
n2 <- 10

data <- list("k1", "k2", "n1", "n2") # to be passed on to JAGS
myinits <-  list(
  list(theta = 0.5))

# parameters to be monitored:	
parameters <- c("theta", "postpredk1", "postpredk2")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file ="Rate_5.txt", n.chains=1, n.iter=1000, 
         n.burnin=1, n.thin=1, DIC=T)
df<- data.frame(samples$BUGSoutput$sims.list$theta)
names(df)[1]<-'theta'
df$postpredk1 <- samples$BUGSoutput$sims.list$postpredk1
df$postpredk2 <- samples$BUGSoutput$sims.list$postpredk2
 	 			 	 			
#ggplot(df, aes(x=theta)) + 
#  geom_density()+
##  theme_bw()+
#  xlim(0,1)


df0<-melt(table(df[,2:3]))

quartz()
ggplot(df0, aes(x=postpredk1, y=postpredk2)) + 
  geom_tile(aes(fill = value), colour = "white") + 
  scale_fill_gradient(low = "white", high = "steelblue")+
  theme_bw()
```

### 3.6 Joint distribution

```{r}
nmax <- 500
k    <- c(16,18,22,25,27)
#k    <- c(16,18,22,25,28)
m    <- length(k)

data <- list("nmax", "k", "m") # to be passed on to JAGS
myinits <-  list(
  list(theta = 0.5, n = nmax/2))

# parameters to be monitored:	
parameters <- c("theta", "n")

# The following command calls JAGS with specific options.
# For a detailed description see the R2jags documentation.
samples <- jags(data, inits=myinits, parameters,
	 			 model.file ="Survey.txt", n.chains=1, n.iter=5000, 
         n.burnin=1, n.thin=1, DIC=T)
# Now the values for the monitored parameters are in the "samples" object, 
# ready for inspection.

d <- data.frame(samples$BUGSoutput$sims.list$theta)
names(d)[1]<-'theta'
d$n <- samples$BUGSoutput$sims.list$n 

## First calculate MLE:
cc <- -Inf
ind <- 0

for (i in 1:length(d$n))
{
	logL <- 0
	for(j in 1:m)
	{		
		logL <- logL+lgamma(d$n[i]+1)-lgamma(k[j]+1)-lgamma(d$n[i]-k[j]+1)
		logL <- logL+k[j]*log(d$theta[i])+(d$n[i]-k[j])*log(1-d$theta[i])
	}
	if (logL>cc) 
  {
		ind <- i
		cc <- logL
	}
}

expct_n = mean(d$n)
expct_theta = mean(d$theta)

a<-ggplot(d, aes(x=n, y=theta)) + 
  geom_point()+
  geom_point(aes(x=expct_n,y=expct_theta),size =6,color='red')+
  geom_point(aes(x=n[ind],y=theta[ind]),size =6,color='blue')+
  theme_bw()

b<-ggplot(d, aes(x=n))+
  geom_histogram()+
  theme_bw()

c<-ggplot(d, aes(x=theta))+
  geom_histogram(binwidth=0.025)+
  theme_bw()+
  coord_flip()+
  scale_y_reverse()

blankPanel<-grid.rect(gp=gpar(col="white"))
quartz()
grid.arrange(a,c,b,blankPanel,ncol=2)
```


Red dot is the marginal expected value of the joint posterior (taken from the mean of each of the marginal distributions, no?). Does this make sense?

Blue dot is the MAP/MLE of the joint posterior (approximation to the mode). Does this make sense?

Note: How much do the likelihoods of the MAP change from the two data sets?

3.6.3 The mode (MLE) is very sensitive to the data. The expected value is not as much.
```{r 3.6.4}
traceplot(samples)
# samples "sweep" through high and low values showning dependence of samples on previous samples
# not ideal! ideally independent


samples_thin <- jags(data, inits=myinits, parameters,
     		 model.file ="Survey.txt", n.chains=1, n.iter=50000, 
         n.burnin=1, n.thin=100, DIC=T)

traceplot(samples_thin)


# with more than 1 chain


myinits <-  list(
  list(theta = 0.5, n = nmax/2), list(theta=0.5, n=nmax/2))
samples_thin <- jags(data, inits=myinits, parameters,
   			 model.file ="Survey.txt", n.chains=2, n.iter=50000, 
         n.burnin=1, n.thin=100, DIC=T)

traceplot(samples_thin)
```
