---
title: "LW Chapter 3"
author: "mht"
date: "October 5, 2014"
output: html_document
---

### 3.1 Inferring a rate

```{r setup}
setwd("~/Repos/psych201s/practicums/")
library(rwebppl)
library(ggmcmc)
#library(gridExtra)
rm(list=ls())
```

```{r}
myFirstModel <- '
// Unpack data
var k = observed_data["k"] // number of heads
var n = observed_data["n"] // number of flips

var model = function() {
   var p = uniform( {a:0, b:1} )    // sample a coin weight from a uniform distribution
   observe({
      data : k,           // Observed k number of Heads
      link: Binomial( {p : p, n: n }) // assuming a Binomial distribution
   })
   return {p: p}
}
'
# k = number of heads
# n = number of coin flips

runModel = function(model, data, numSamples = 10000, method = "MCMC") {
  # The following command calls WebPPL with specific options.
  wp <- webppl(
    program_code = model,
    data = data,
    data_var = "observed_data",
    inference_opts = list(method=method, samples = numSamples, burn = numSamples/10),
    ggmcmc = TRUE,
    model_var = "model",
    packages = c("./utils")
    )
}
```

Writing a model as a string in R is convenient, but it is also convenient to see syntax-highlighting (like you get syntax highlighting for R). To get this, we can just write out model as a .wppl file in our folder. You can open the file in RMarkdown so you can view it and edit it here. In the bottom-right of the editor, you can change the language to JavaScript to get some helpful comments. (Note there are a few differences between JavaScript and WebPPL so all of the comments may not apply).

```{r}
observed_data <- list(k = 5, n = 10) 
runModel(myFirstModel, observed_data) %>%
  ggs_histogram()
```


#3.1.1
Hint: remember that the probability of an interval is area under the curve


#3.1.2


#3.1.3



```{r, echo=FALSE}
# Enter new k & n values here:
observed_data <- list(k = ..., n = ...) 
runModel(myFirstModel, observed_data)
```


#3.1.4

Hint: we used 10000 samples above

```{r}
observed_data <- list(k = 5, n = 10) 
runModel(myFirstModel, observed_data, numSamples = ...)

observed_data <- list(k = 50, n = 100) 
runModel(myFirstModel, observed_data, numSamples = ...)
```


#3.1.5

```{r}
observed_data <- list(k = ..., n = ...) 
runModel(myFirstModel, observed_data)
```

#3.1.6

```{r}
observed_data <- list(k = ..., n = ...) 
runModel(myFirstModel, observed_data)
```
  
### 3.2 Difference between two rates

```{r}
# k1 = number of heads for coin 1
# k2 = number of heads for coin 2
# n1 = number of coin flips for coin 1
# n2 = number of coin flips for coin 2
differenceInWeightsModel <- '
// Unpack data
var k1 = observed_data["k1"]
var k2 = observed_data["k2"]
var n1 = observed_data["n1"]
var n2 = observed_data["n2"]

var model = function() {
  // sample two rates from uniform priors
  var p1 = uniform({a: 0, b: 1})
  var p2 = uniform({a: 0, b: 1})

  // Take first observation into account
  observe({
    data : k1,          
    link: Binomial( {p : p1, n: n1 }) 
  }) 

  // Take second observation into account
  observe({
    data : k2,          
    link: Binomial( {p : p2, n: n2 }) 
  }) 

  // Return the difference
  return {delta : p1 - p2}
}
'
```

Run the model:

```{r}
observed_data = list(k1 = 5, n1 = 10, k2 = 7, n2 = 10)
runModel(differenceInWeightsModel, observed_data) %>%
  ggs_histogram()
```

# 3.2.1

Run with first set of params:

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(differenceInWeightsModel, observed_data) %>%
  ggs_histogram()
```

With second set of params:

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(differenceInWeightsModel, observed_data) %>%
  ggs_histogram()
```

# 3.2.2

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(differenceInWeightsModel, observed_data) %>%
  ggs_histogram() 
```

# 3.2.3



### 3.3 Inferring a common rate

```{r}
commonRateModel <- '
// Unpack data
var k1 = observed_data["k1"]
var k2 = observed_data["k2"]
var n1 = observed_data["n1"]
var n2 = observed_data["n2"]

var model = function() {
  // Sample rate from uniform distribution
  var p = uniform( {a:0, b:1} )
  
  // account for first data point using p
  observe({
    data : k1,           // Observed k number of Heads
    link: Binomial( {p : p, n: n1 }) // assuming a Binomial distribution
  }) 

  // account for second data point also using p
  observe({
    data : k2,           // Observed k number of Heads
    link: Binomial( {p : p, n: n2 }) // assuming a Binomial distribution
  }) 

  return {p : p}
}
'
```

Run the model, example:

```{r}
observed_data = list(k1 = 5, n1 = 10, k2 = 7, n2 = 10)
runModel(commonRateModel, observed_data) 
  ggs_density() + xlim(0,1)
```

# 3.3.1

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggs_density() + xlim(0,1)
```

# 3.3.2

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggs_density() + xlim(0,1)
```

# 3.3.3

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggs_density() + xlim(0,1)
```

```{r}
observed_data = list(k1 = ..., n1 = ..., k2 = ..., n2 = ...)
runModel(commonRateModel, observed_data) %>%
  ggs_density() + xlim(0,1)
```

### 3.4 Prior and posterior prediction

** This example is broken due to issue #20 in rwebppl **

```{r}
firstModelWithPredictives <- '
// Unpack data
var k = observed_data["k"] // number of heads
var n = observed_data["n"] // number of flips

var model = function() {
   var p = uniform( {a:0, b:1} )    // sample a coin weight from a uniform distribution
   
   observe({
      data : k,           // Observed k number of Heads
      link: Binomial( {p : p, n: n }) // assuming a Binomial distribution
   }) 

   var posteriorPredictive = sample(Binomial({p : p, n: n}))
   var prior = uniform({ a: 0, b: 1});
   var priorPredictive = sample(Binomial({p : prior, n: n}))
   return {prior: prior, 
           priorPredictive : priorPredictive,
           posterior : p,
           posteriorPredictive : posteriorPredictive};
}
'
```


```{r}
observed_data <- list(k = 1, n = 15)
runModel(firstModelWithPredictives, observed_data) %>%
  ggs_density()
```

# 3.4.1


# 3.4.2

```{r}
newPriorModel <- '
// Unpack data
var k = observed_data["k"] // number of heads
var n = observed_data["n"] // number of flips

var model = function() {
   var p = beta({a : ..., b : ...})  // TRY NEW PRIOR HERE
   
   observe({
      data : k,           // Observed k number of Heads
      link: Binomial( {p : p, n: n }) // assuming a Binomial distribution
   }) 

   var posteriorPredictive = sample(Binomial({p : p, n: n}))
   var prior = uniform({ a: 0, b: 1});
   var priorPredictive = sample(Binomial({p : prior, n: n}))
   return {prior: prior, 
           priorPredictive : priorPredictive,
           posterior : p,
           posteriorPredictive : posteriorPredictive};
}
'
```

# 3.4.3

```{r}
newPredictiveModel <- '
// Unpack data
var k = observed_data["k"] 
var n = observed_data["n"] 
var n_prime = ... // ENTER NEW N_PRIME FOR PREDICTIVES

var model = function() {
   var p = beta({a : 1, b : 1})  
   
   observe({
      data : k,          
      link: Binomial( {p : p, n: n }) 
   }) 

   // AFTER LEARNING ABOUT P, WE CAN MAKE PREDICTIONS FOR N_PRIME
   var posteriorPredictive = sample(Binomial({p : p, n: n_prime }))
   var prior = uniform({ a: 0, b: 1});
   var priorPredictive = sample(Binomial({p : prior, n: n}))
   return {prior: prior, 
           priorPredictive : priorPredictive,
           posterior : p,
           posteriorPredictive : posteriorPredictive};
}
'
```

# 3.4.4

```{r}
TrompetterModel <- '
// Original data
var k = 24
var n = 121

// Uncomment to try your own data!
// var k = observed_data["k"] 
// var n = observed_data["n"] 

var model = function() {
   var p = beta({a : 1, b : 1})  
   
   observe({
      data : k,          
      link: Binomial( {p : p, n: n }) 
   }) 

   var posteriorPredictive = sample(Binomial({p : p, n: n}))
   var prior = uniform({ a: 0, b: 1});
   var priorPredictive = sample(Binomial({p : prior, n: n}))
   return {prior: prior, 
           priorPredictive : priorPredictive,
           posterior : p,
           posteriorPredictive : posteriorPredictive};
}
'
```

### 3.5 Posterior Prediction

** This example is also broken due to issue #20 in rwebppl **

```{r}
commonRateModelWithPredictives <- '
// Unpack data
var k1 = observed_data["k1"]
var k2 = observed_data["k2"]
var n1 = observed_data["n1"]
var n2 = observed_data["n2"]

var model = function() {
  // Sample rate from uniform distribution
  var p = uniform( {a:0, b:1} )
  
  // account for first data point using p
  observe({
    data : k1,           // Observed k number of Heads
    link: Binomial( {p : p, n: n1 }) // assuming a Binomial distribution
  }) 

  // account for second data point also using p
  observe({
    data : k2,           // Observed k number of Heads
    link: Binomial( {p : p, n: n2 }) // assuming a Binomial distribution
  }) 
  var posteriorPredictive = {k1 : sample(Binomial({p : p, n: n1})),
                             k2 : sample(Binomial({p : p, n: n2}))}
  return {p : p,
          posteriorPredictive : posteriorPredictive}
}
'
```

```{r}
observed_data <- list(k1 = 0, n1 = 10, k2 = 10, n2 = 10)
runModel(commonRateModelWithPredictives, observed_data)
```


# 3.5.1

# 3.5.2

# 3.5.3

### 3.6 Joint distributions

```{r}
# k: List of m success counts
Survey <- '
// Unpack data
var nmax = observed_data["nmax"]
var k = observed_data["k"]
var probs = repeat(nmax, function() {return 1/nmax;});
var vals = _.range(1, nmax + 1)

// Inferring a Rate
var model = function() {
  var n = categorical( {ps: probs, vs: vals} );
  var p = beta({a: 1, b: 1})
  observe({
    data : k,           // Observed k number of Heads
    link: Binomial( {p : p, n: n }) // assuming a Binomial distribution
  })
  return {n: n, p: p}
}
'
```

```{r}
observed_data <- list(k = c(16, 18, 22, 25, 27), nmax = 500)
output <- runModel(Survey, observed_data, numSamples = 100000, method = "incrementalMH") 
ggs_pairs(output)
```

# 3.6.1

# 3.6.2

# 3.6.3

```{r}
observed_data <- list(k = c(16, 18, 22, 25, 28), nmax = 500)
output <- runModel(Survey, observed_data, numSamples = 100000, method = "incrementalMH") 
ggs_pairs(output)
```

# 3.6.4

** Can't do this in rwebppl right now; need order of samples ** 
```{r}
observed_data <- list(k = c(16, 18, 22, 25, 27), nmax = 500)
output <- runModel(Survey, observed_data, numSamples = 100000, method = "incrementalMH") 
ggs_pairs(output)
```