---
title: 'Psych 201s: Distributions and functional programming in WebPPL'
author: "mht"
date: "June 6, 2016"
output: html_document
---

R has built-in functions for sampling from distributions and looking up probabilities. 
These kinds of distributions can encode our "prior beliefs" about the outcome of an experiment. 
But we will want to update these beliefs with our observed data.
For that, we won't be able to use just base R.

We will switch to using WebPPL, which is what's called a *probabilistic programming language* (PPL). PPLs are languages for doing Bayesian inference. WebPPL is also also a *functional* programming language, so we will go over some basic ideas of functional programming here as well. Functional programming languages are everywhere: R is a functional language, JavaScript is too.

We'll introduce WebPPL first with distributions (like we did in R).
A full list of distributions in WebPPL can be found [here](http://webppl.readthedocs.io/en/master/distributions.html)

# Bernoulli distribution

The simplest probability distribution results from the random process of flipping a coin.
It is called a Bernoulli distribution (some people just call it a flip).

In WebPPL, you can access the Bernoulli distribution using the call `Bernoulli`

```{r Bernoulli}
library(rwebppl)
webppl("Bernoulli({ p: 0.5 })")
```

Using RWebPPL, we will be writing the WebPPL model as a string `"Bernoulli({ p: 0.5 })"`.
The `Bernoulli` distribution takes a single parameter `p` inside of a JavaScript style object `{p: 0.5}` 
(Technical note: An object is the exactly the idea of a dictionary from Python, or a JSON object from JavaScript; it's very similar conceptually to a dataframe in R).

What happens when we call `Bernoulli`? (Note: the capital letter is important).

We get out a table of probabilities. 
Each possible outcome of the random process (here, a coin flip) is an element of the **support**. 
Associated with each element of a support is the probability of that outcome. 

Let's visualize this.

```{r Bernoulli.viz}
d <- webppl("Bernoulli( { p: 0.5 } )")

ggplot(d, aes(x = "Bernoulli" , fill = support, y = prob ))+
  geom_bar(stat='identity', width = 0.5)
```

Try changing around p. What happens? If TRUE / FALSE are two responses in a 2AFC, what does p represent?

Distributions in WebPPL can be sampled from. That is, you can flip the coin.

```{r flip}
webppl("sample( Bernoulli({ p: 0.5 }) )")
```

You might notice this call is appreciably slower than `rbinom` because we are calling out to the language WebPPL. 
This is a fixed-cost, however, and will be worth it once we start considering more interesting cases than just a single coin flip.

Note, unlike `Bernoulli` (which can returns the same thing each time you call it), 
`sample(Bernoulli(...))` is **stochastic**: It is a random sample from the Bernoulli distribution.
For shorthand, you can sample from a distribution using the lower-case name (e.g. `bernoulli({p:0.5})`); you can even drop the parameter object and just give it the parameter value (i.e., `bernoulli(0.5)`). 
However, we will continue to use the longer `sample(Distribution ( ... ) )` notation for explicitness.

# Binomial distribution

What about if we wanted to flip the coin multiple times? (or, equivalently, what if we collected some number of participants worth of data on a 2AFC).

If we don't don't care about the order of the flips (or, the identities of our participants), then the data is distributed according to what's called the Binomial distribution.

```{r binomial.sample}
webppl("sample( Binomial({n:3, p:0.5 }) )")
```

`Binomial` takes in parameters `n` and `p`, bundled together in the same object structure we saw above. `p` is the weight of the coin; same as in `bernoulli`. `n` is the number of times you flip it (or, the number of identical coins you flip).
Note that we know are getting out a number: it is the number of coins that came up heads (or, `TRUE`, as we saw above).

What does the distribution look like?

```{r binomial.dist}
webppl("Binomial( {n:3, p:0.5 } )")
```

Let's visualize this.

```{r binomial.dist.viz}
webppl("Binomial( {n:3, p:0.5 } )") %>%
  ggplot(., aes( x = support, y = prob ))+
  geom_bar( stat = 'identity', position = position_dodge() )
```

Interesting. So 1 and 2 heads are more probable than 0 or 3. Do you understand why?

Try changing `n`. What happens? Why?  ...

Try changing `p`. What happens? Why?  ...

Try changing `n` to 6. And then play around with `p`. Imagine each support element is a position on a likert scale (add 1 to all the values if it makes you happy). What might `p` represent?

# Sampling using functional programming

Often times, we don't have the true distribution to visualize. In these cases, we will have **samples** from the true distribution. We can use the function `repeat` to repeatedly sample. Check it out:

```{r sample.repeat}
webppl("
  var sampleLikert = function () { return sample( Binomial({n:6, p:0.8}) ) + 1 }
  repeat(10, sampleLikert)
")
```

Now inside of our string `"..."`, we've actually written a whole program! 
In WebPPL, when you declare a new variable, you must write `var varName = ...`
On the first line, we have made a new variable called `sampleLikert`. 
`sampleLikert` is a function (you can tell because it says `= function`). 
The empty parentheses `()` is where the **arguments** of the function would normally go, but in this case, we've made a function that doesn't take any arguments. 
Inside of the curly-braces `{}` is what the function does. 
Here, what it does is `sample` from a `Binomial` distribution and adds 1 to it. 

This procedure is called *function abstraction*. 
We've taken a process that we want the computer to do (sample from a binomial), and packaged it up into a function `sampleLikert`, which can later be called. 
This is a basic idea in functional programming.

Rather than call it, however, we pass it to another function: `repeat`, which ends up calling it 10 times.

Notice that we can't actually call `repeat` on `sample(Binomial( {n:6, p:0.8} ) )`. The reason is that `sample(Binomial({n:6, p:0.8}))` is not a function (it is a *call* to a function, and is actually a sample), and `repeat` wants to repeat **a function**. So what we did here was make a function called `sampleLikert`, and tell repeat to repeat that. 

Technically, we didn't have to make a new function and give it a name. We could have passed `function()...` directly into repeat.

Let's write it that way, and capture the output and visualize it. We'll try it with 1000 samples.

```{r sample.repeat.2}
x <- webppl("
  repeat(1000, function () { return sample(Binomial( {n:6, p:0.8} )) + 1 } )
")
print(x)
qplot(x)
```

**Challenge problem**: Make a function that samples from 2 Binomials (e.g. with n = 6) and adds them together.
Then, repeat the function 1000 times and visualize it.

```{r}
x <- webppl("
  var twoBinomials = ...
  repeat(1000, twoBinomials )
")

qplot(x)

```


# Gaussian distribution

Let's take a look at a Gaussian distribution.

A Gaussian has 2 parameters: mean `mu` and standard deviation `sigma` (careful: in some systems, the Gaussian is parameterized by the mean and *variance*... *variance* being the standard deviation squared).

```{r gauss}
webppl("Gaussian( {mu: 0, sigma:1 } ) ")
```

In this situation, WebPPL doesn't give us a pretty probability table. Why not?

The answer has to do with the **support** of the Gaussian distribution. Whereas before, we had a *finite* support (all of the possible outcomes of a coin flip, or several coin flips). The support of a Gaussian is all numbers.. it is *infinite*. That is, every number has some probability under a Gaussian distribution (it may be a very small probability).
We can't print out a probability table because the table would be infinitely long!

You may think this is a nit-picky detail, and that we could make some discretization or "binning" to show the distribution. You are right, we could do that. How could we do that?
Well, we could look up the probability for several (but not infinite) different values. 
In WebPPL, Distributions come with this ability. We're going to use the function `probability` to look up the probability of a number under a `Gaussian` distribution.
This function is not in the basic WebPPL language, but is in a package we've made for your convenience called `utils`.

```{r gauss.prob}
webppl("
  var x = 0
  probability(x, Gaussian({mu: 0, sigma:1 }))
", packages = c("./utils"))
```

Let's do this for several x-values. 
To do this, we're learn another basic trick from functional programming.
If we want to do the same procedure (i.e., apply the same function, e.g. call `probability(..)`) many times to different input values, we can use a  function called `map`.
`map` takes in 2 arguments: a function, and a list of values to apply that function to.
It returns to you a list of values after the function has been applied. (It's sort of like a for-loop, but not really.)

Because we are passing actual values, we want to make a function that *takes in an argument* (as opposed to repeat, where we made a function with no arguments).

```{r gauss.discretized}
ourWebPPLprogram <- "
  var gaussianProbability = function(x) { return probability(x, Gaussian( {mu: 0, sigma:1 } )) }
  map(
    gaussianProbability, // first argument is a function
    [-5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5] // second argument is a list of values (each of these gets substituted for x in gaussianProbability)
  )
"

discreteGaussian <- webppl(ourWebPPLprogram, packages = c("./utils"))

print(discreteGaussian)

data.frame(
  bins = c(-5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5),
  probs = discreteGaussian
) %>%
ggplot(., aes( x = bins, y = probs ) )+
  geom_bar( stat = 'identity', position = position_dodge() )
```

Try varying `mu` and `sigma`.

Now let's make a continuous distribution from samples.

```{r}
webppl("
repeat(
  10000,
  function () { return sample( Gaussian( {mu: 0, sigma:1 } ) ) }
)
", packages = c("./utils")) %>% 
  qplot()+
  xlim(-5,5)
```



