---
title: 'Psych 201s: Distributions using WebPPL'
author: "mht"
date: "June 6, 2016"
output: html_document
---

A fundamental concept in probability and statistics is that of a probability distribution.
There are different "families" of distributions, corresponding to different "random processes". 

We'll consider a few here. 
We'll explore the distributions using WebPPL.

# Bernoulli distribution

The simplest probability distribution results from the random process of flipping a coin.


My favorite function is the random sampling function. 
syntax: rbinom(n = number_of_experiments, size = number_of_trials, prob = probability_of_success)
for a single coin flip: number_of_trials = 1
for a fair coin, probability_of_success = 0.5 (but for a biased coin, it could be between 0 and 1)
outcome is either 0 (tails) or 1 (heads)...

```{r samples}
rbinom(n = 1, size = 1, prob = 0.5)
```

In WebPPL, you can access the flip of a single coin using the primitive `flip`
```{r flip}
library(rwebppl)
webppl("bernoulli( { p : 0.5 } )")
```

Using RWebPPL, we will be writing the WebPPL model as a string `"flip({p:0.5})"`.
`flip` takes a single parameter `p` inside of a parameter object `{p:0.5}` (Note: This is the exactly the idea of a dictionary from Python, or a JSON object from JavaScript).

The function call is appreciably slower than `rbinom` because we are calling out to language WebPPL. 
This call is a fixed-cost, however, and will be worth it once we start considering more interesting cases than just a single coin flip.

Here, we've collected one sample. If we collect a bunch of samples, it would start to approximate the true distribution.

In WebPPL, distributions can be accessed directly. They are denoted by their proper names with capital letters. A full list of distributions can be found [here](http://webppl.readthedocs.io/en/master/distributions.html)

What happens if we make the same sort of call on the distribution `Bernoulli`?
```{r Bernoulli}
webppl("Bernoulli( { p: 0.5 } )")
```

We get out a table of probabilities. Each possible outcome of the random process (here, a coin flip) is an element of the **support**. Associated with each element of a support is the probability of that outcome. Note, unlike `bernoulli` (which can return different values each time you call it), `Bernoulli` is deterministic: It is a reflection of the true distribution. 

# Binomial distribution

What about if we wanted to flip the coin multiple times?
If we don't don't care about the order of the flips, then this is what's called the Binomial distribution.

```{r binomial.sample}
webppl("binomial( {n:3, p:0.5 } )")
```

`binomial` takes in parameters `n` and `p`, bundled together in the same object structure we saw above. `p` is the weight of the coin; same as in `bernoulli`. `n` is the number of times you flip it (or, the number of coins you flip).
Note that we know are getting out a number: it is the number of coins that came up heads (or, `TRUE`, as we saw above).

What does the distribution look like?

```{r binomial.dist}
webppl("Binomial( {n:3, p:0.5 } )")
```

Interesting. So 1 and 2 heads are more probable than 0 or 3. Do you understand why?

### Probability density

dbinom gives you the probability of a certain outcome (technically, this is called the *probability density*)
syntax: dbinom(x=outcome, size=number_of_trials, prob=probability_of_success) [also see ?dbinom]

```{r density}
dbinom(x=0, size=1, prob=0.5)
# you can also supply a vector of outcomes
dbinom(x=c(0,1), size=1, prob=0.9)
```

What happens if you set the outcome to a number other than 0 or 1?
What happens if you set the probability_of_success to a number less than 0 or greater than 1?

Let's not explore the number_of_trials parameter yet.

### Quantile function

qbinom is the opposite of dbinom.
dbinom: outcome -> probability
qbinom: probability -> outcome

It's not exactly true but it's close. Actually, qbinom will give you the max(outcome).

```{r quantile}
qbinom(0.8, size=1, prob=0.5)
```

### Cumulative probability

pbinom gives you the probability of outcome<=q (equal or less than)
For the bernoulli distribution, this isn't very interesting as there are only two outcomes (0 or 1). 

```{r cumulative}
pbinom(q = 0, size=1, prob= 0.5)
```


