---
title: 'Psych 201s: Distributions'
author: "mht"
date: "May 5, 2016"
output: html_document
---

A fundamental concept in probability and statistics is that of a probability distribution.
There are different "families" of distributions, corresponding to different "random processes". 

We'll consider a few here. 
First, we'll explore the distributions using R. 
Then, we'll see how to look at them using JAGS.

# Bernoulli distribution

The simplest probability distribution results from the random process of flipping a coin.

There are 4 functions in R you can use to access probility distributions.

### Samples

My favorite function is the random sampling function. 
syntax: rbinom(n = number_of_experiments, size = number_of_trials, prob = probability_of_success)
for a single coin flip: number_of_trials = 1
for a fair coin, probability_of_success = 0.5 (but for a biased coin, it could be between 0 and 1)
outcome is either 0 (tails) or 1 (heads)...

```{r samples}
rbinom(100, size=1, prob=0.5)

# kind of messy just looking at numbers, pass it qplot [this may require the tidyr package]
rbinom(100, size=1, prob=0.5) %>%
  qplot()

# note: this is the same as qplot(rbinom(100, size=1, prob=0.5)) 
# qplot is smart and if you give it a list of numbers, it will make you a histogram automagically
```

In all honesty, to think about distributions, I use the rbinom function 98% of the time. 
But I'll introduce the other functions here, just for completeness.

### Probability density

dbinom gives you the probability of a certain outcome (technically, this is called the *probability density*)
syntax: dbinom(x=outcome, size=number_of_trials, prob=probability_of_success) [also see ?dbinom]

```{r density}
dbinom(x=0, size=1, prob=0.5)
# you can also supply a vector of outcomes
dbinom(x=c(0,1), size=1, prob=0.9)
```

What happens if you set the outcome to a number other than 0 or 1?
What happens if you set the probability_of_success to a number less than 0 or greater than 1?

Let's not explore the number_of_trials parameter yet.

### Quantile function

qbinom is the opposite of dbinom.
dbinom: outcome -> probability
qbinom: probability -> outcome

It's not exactly true but it's close. Actually, qbinom will give you the max(outcome).

```{r quantile}
qbinom(0.8, size=1, prob=0.5)
```

### Cumulative probability

pbinom gives you the probability of outcome<=q (equal or less than)
For the bernoulli distribution, this isn't very interesting as there are only two outcomes (0 or 1). 

```{r cumulative}
pbinom(q = 0, size=1, prob= 0.5)
```


